---
title: "R package primefactr"
author: "Florian Privé"
date: "August 10, 2016"
tags: R
output: 
  html_document:
    toc: true
---

***

In this post, I will present my first R package. It makes use of [Prime Factorization](https://en.wikipedia.org/wiki/Prime_factor) for computations.

This small R package was initially developed to compute [hypergeometric probabilities](https://en.wikipedia.org/wiki/Hypergeometric_distribution) which are used in Fisher's exact test, for instance. It was also a way to get introduced with CRAN submission :').

## Installation and Attachment

```{r, eval=FALSE}
## Installation
install.packages("primefactr")
```
```{r, eval=TRUE}
## Attachment
library("primefactr")
```

## Features

### Main feature
For instance, to compute $$P(X = k) = \dfrac{\binom{K}{k}~\binom{N-K}{n-k}}{\binom{N}{n}} = \dfrac{K!~(N-K)!~n!~(N-n)!}{k!~(K-k)!~(n-k)!~(N-K-n+k)!~N!},$$ you can use 
```{r}
f <- function(k, N, K, n) {
  ComputeDivFact(c(K, (N-K), n, (N-n)),
                 c(k, (K-k), (n-k), (N-K-n+k), N))
}
f(4, 50, 5, 10)
f(5, 50, 5, 10)
```
You can check the results [here](https://en.wikipedia.org/wiki/Hypergeometric_distribution#Application_and_example).

Let us now check large numbers:
```{r}
f(k = 1000, N = 15100, K = 5000, n = 3100)
```
A direct approach would require computing `factorial(15100)`, while `factorial(100) = 9.332622e+157`.


### Implementation

This uses a Prime Factorization to simplify computations.

I code a number as follows,
$$number = \prod i^{code[i]},$$
or, which is equivalent,
$$\log(number) = \sum code[i] \times \log(i).$$
For example, 

- $5$ is coded as (0, 0, 0, 0, 1),
- $5!$ is coded as (1, 1, 1, 1, 1),
- $8!$ is coded as (1, 1, 1, 1, 1, 1, 1, 1).

So, to compute $8! / 5!$, you just have to substract the code of $5!$
from the code of $8!$ which gives you (0, 0, 0, 0, 0, 1, 1, 1).

Then there is the step of Prime Factorization:

Factorization by 2:

- it becomes (0, 2, 1, 1, 0, 0, 1, 0) because $8 = 4 \times 2$ and $6 = 3 \times 2$,
- then it becomes (0, 4, 1, 0, 0, 0, 1, 0) because $4 = 2^2$.

This is already finished (this is a small example). You get that $8! / 5! = 2^4 \times 3^1 \times 7^1$. Let us verify:
```{r}
cat(sprintf("%s == %s", factorial(8) / factorial(5), 2^4 * 3 * 7))
```

### Play with primes

You can also test if a number is a prime and get all prime numbers up to a certain number.

## Submission to CRAN

It was easier than I thought.
I've just followed the instructions of the book [R packages](http://r-pkgs.had.co.nz/) by Hadley Wickham. 
I had two notes:

1. It is my first submission.
2. File README.md cannot be checked without ‘pandoc’ being installed.
For this note, I used the same comment as [here](https://github.com/klarsen1/Information/blob/master/cran-comments.md) and CRAN didn't complain.
