---
title: "Getting My Colleagues Hooked on R"
author: "Florian Privé"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    css: pResentation-styles.css
    self_contained: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  warning = FALSE, message = FALSE, 
  fig.align = 'center', comment = "")
```

```{r, include=FALSE}
pacman::p_load(googlesheets, gsubfn, stringr)
printf <- function(...) cat(sprintf(...))

sheet <- gs_key("113h74NVW_VZWuF9EttiYTNVONS-SXEvi7MwGizZD2LE")

sheet.content <- gs_read(sheet)

responses <- sheet.content[, 2]
n <- nrow(responses)
```


## What `r n` of you wanted to learn

```{r, echo=FALSE, results='asis'}
url <- "https://docs.google.com/forms/d/13Hrr9x-3ASaPKPJGRpKzsGWlkwu59LGITtw1ldG7FAU/edit#responses"
url <- url(url)
url.lines <- readLines(url, encoding = "UTF-8", warn = FALSE)
pattern <-  "\\[\"([^\"]*)\",,,,0\\]"
questions <- unlist(gsubfn::strapply(url.lines, pattern))

counts <- str_count(responses, coll(questions))

counts.lvl <- setdiff(sort(unique(counts), decreasing = TRUE), 0)

for (n in counts.lvl) {
  if (n == 2) printf("\n***\n")
  printf("- for **%d** of you:\n", n)
  q.tmp <- questions[counts == n]
  for (q in q.tmp) {
    printf("    - %s\n", q)
  }
}
```

<!-- ## What is R?  -->

<!-- <iframe width="640" height="360" src="https://www.youtube.com/embed/TR2bHSJ_eck" frameborder="0" allowfullscreen></iframe> -->


## Some facts about the growth of R:

- R is #5 of all programming languages ([IEEE Spectrum, July 2016](https://www.r-bloggers.com/r-moves-up-to-5th-place-in-ieee-language-rankings/))

```{r, echo=FALSE}
knitr::include_graphics("http://revolution-computing.typepad.com/.a/6a010534b1db25970b01bb092485d1970d-800wi")
```

<!-- The other four languages in the top 5 (C, Java, Python amd C++) are all general-purpose languages, suitable for just about any programming task. R by contrast is a language specifically for data science, and its high ranking here reflects both the critical importance of data science as a discipline today, and of R as the language of choice for data scientists. -->

---

```{r, echo = FALSE}
thepage <- readLines('https://cran.r-project.org/web/packages/')
theline <- grep("Currently, the CRAN package repository features [0-9]+ available packages.", thepage)
thesplit <- strsplit(thepage[theline], split = " ")[[1]]
thepos <- match("available", thesplit) - 1
n <- thesplit[thepos]
```

- There are now  `r n` available packages on CRAN ([CRAN: Contributed Packages, `r Sys.Date()`](https://cran.r-project.org/web/packages/))

```{r, echo=FALSE, out.height=500}
knitr::include_graphics("http://a3.typepad.com/6a017d41eeee1a970c01bb08ef2103970d-pi")
```

---

- There are many R conferences:
    - useR!: 900+ people in 2016,
    - eRum: european R users meeting,
    - EARL: many people from the Industry,
    - Rencontres R: Grenoble in 2015,
    - satRdays,
    - R/Finance & R in Insurance.
    
- The R blogosphere is huge: [R-bloggers](https://www.r-bloggers.com/) has
    - nearly 600 bloggers,
    - 36K followers on Twitter,
    - 39K on Facebook,
    - very interesting posts every day!

## Manipulating data? Ask Hadley Wickham!

R packages that he has developped (from [his website](http://hadley.nz/)):

- Data science
    - <a href="http://ggplot2.org">ggplot2</a> for visualising data.
    - <a href="http://github.com/hadley/dplyr">dplyr</a> for manipulating data.
    - <a href="http://github.com/hadley/tidyr">tidyr</a> for tidying data.
    - <a href="http://github.com/hadley/stringr">stringr</a> for working with strings.
    - <a href="http://github.com/hadley/lubridate">lubridate</a> for working with date/times.
    
---

- Data import
    - <a href="http://github.com/hadley/readr">readr</a> for reading .csv and fwf files.
    - <a href="http://github.com/hadley/readxl">readxl</a> for reading .xls and .xlsx files.
    - <a href="http://github.com/hadley/haven">haven</a> for SAS, SPSS, and Stata files.
    - <a href="http://github.com/hadley/httr">httr</a> for talking to web APIs.
    - <a href="http://github.com/hadley/rvest">rvest</a> for scraping websites.
    - <a href="http://github.com/hadley/xml2">xml2</a> for importing XML files.
    
- Software engineering
    - <a href="http://github.com/hadley/devtools">devtools</a> for general package development.
    - <a href="http://github.com/klutometis/roxygen">roxygen2</a> for in-line documentation.
    - <a href="http://github.com/hadley/testthat">testthat</a> for unit testing.

## Introduction to dplyr (from its vignette)

```{r, collapse=TRUE}
library(nycflights13)
dim(flights)
head(flights)
```


***

```{r}
require(dplyr)
```


Dplyr aims to provide a function for each basic verb of data manipulation:

- ``filter()`` (and ``slice()``)
- ``arrange()``
- ``select()`` (and ``rename()``)
- ``distinct()``
- ``mutate()`` (and ``transmute()``)
- ``summarise()``
- ``sample_n()`` (and ``sample_frac()``)

*** 

```{r}
filter(flights, month == 1, day == 1)
```

***

```{r}
arrange(flights, desc(dep_delay))
```

***

```{r}
mutate(flights, gain = arr_delay - dep_delay,
  speed = distance / air_time * 60)
```

***

```{r}
flights2 <- flights %>%
  filter(month == 1, day == 1) %>%
  arrange(desc(dep_delay)) %>%
  mutate(gain = arr_delay - dep_delay,
         speed = distance / air_time * 60)
head(flights2)
```




## Elegant visualization tools: [ggplot2](http://ggplot2.org/)

```{r, out.height=380, out.width=600}
require(ggplot2)
p <- qplot(dep_delay, arr_delay, data = flights2, 
      main = "Flights which take off late arrive late. Surprising!")
print(p)
```

## Adding layers

```{r}
p + geom_smooth()
```

## More: go check this book

```{r}
citation("ggplot2")
```


## Some extensions are available [here](https://www.ggplot2-exts.org/)

```{r}
ggExtra::ggMarginal(p, type = "histogram")
```


## [ggmap](https://github.com/dkahle/ggmap): maps with ggplot2

```{r, echo=FALSE}
knitr::include_graphics("http://revolution-computing.typepad.com/.a/6a010534b1db25970b0167689d5031970b-800wi")
```

## Interactive visualizations tools: [plotly](https://plot.ly/ggplot2/)

```{r}
plotly::ggplotly(p)
```
<!-- <center> -->
<!-- ```{r, message=FALSE, out.height=380, out.width=580, echo=FALSE} -->
<!-- library(plotly) -->
<!-- ggplotly(p) -->
<!-- ``` -->
<!-- </center> -->

## Interactive apps: [Shiny](http://shiny.rstudio.com/)

Live demo!

- From the Shiny website
- My own shiny app: `shiny::runGitHub("privefl/repartitions_equipes")`
- A game: [Lights Out](https://daattali.com/shiny/lightsout/)

[Advanced tips and tricks](https://github.com/daattali/advanced-shiny)

## Use of C++ code when needed



## Bigmemory

- On-disk matrices
- types: ``char``, ``short``, ``int``, ``float``, ``double``
- Access with `[i, j]` as a matrix 
- Access via C++ code with `[j][i]`
- Easy use of parallelisation with shared matrices

## Example with foreach and bigmemory

> - Say you have:
    - A SNP big.matrix X stored on-disk in directory _backingfiles_,
    - Infos on the positions of the SNPs (the first 40,000 SNPs are in chromosome 1, then 38,000 are in chromosome 2, etc.),

> - And you have to do some computations which are independent with respect to chromosomes. You want to use __Parallel Computing__!

> - How to do use Parallel Computing on massive genotype matrices?

***

```{r, eval = FALSE, out.height=300}
DO_all <- function(X, infos, ncores) {
  DO_chr <- function(X.desc, lims) {
    X.chr <- sub.big.matrix(X.desc, 
                          firstCol = lims[1], 
                          lastCol = lims[2],
                          backingpath = "backingfiles")
    ## Do something with X.chr (such as imputing)
  }
  range.chr <- LimsChr(infos)
  X.desc <- describe(X)
  obj <- foreach(chr = 1:nrow(range.chr), 
                 .packages = "bigmemory")
  expr_fun <- function(chr) {
    DO_chr(X.desc, range.chr[chr, ])
  }
  res <- foreach2(obj, expr_fun, ncores)
}
```

***

```{r, eval = FALSE}
LimsChr <- function(infos) {
  map.rle <- rle(infos$map$chromosome)
  upper <- cumsum(map.rle$length)
  lower <- c(1, upper[-length(upper)] + 1)
  cbind(lower, upper, "chr" = map.rle$values)
}

foreach2 <- function(obj, expr_fun, ncores) {
  if (is.seq <- (ncores == 1)) {
    foreach::registerDoSEQ()
  } else {
    cl <- parallel::makeCluster(ncores)
    doParallel::registerDoParallel(cl)
  }
  res <- eval(parse(
    text = sprintf("foreach::`%%dopar%%`(obj, expr_fun(%s))",
                   obj$argnames)))
  if (!is.seq) parallel::stopCluster(cl)
  return(res)
}
```

## We have [RStudio](https://www.rstudio.com/)

Live demo!

- Code highlighting/autocompletion
- Help > Cheatsheets
- Panels (Git, ...)
- debugger
- [Notebooks](http://rmarkdown.rstudio.com/r_notebooks.html)

More tips: [RStudio Tips](https://twitter.com/rstudiotips) on Twitter

## Free books to learn about R:

- Advanced R programming: 
    - [Efficient R Programming](https://csgillespie.github.io/efficientR/preface.html)
    - [Advanced R](http://adv-r.had.co.nz/)
- Reporting:
    - [Getting used to R, RStudio, and R Markdown](https://ismayc.github.io/rbasics-book/index.html)
- Data analysis:
    - [R for Data Science](http://r4ds.had.co.nz/)
    - [An Introduction to Statistical Learning, with Applications in R](http://www-bcf.usc.edu/~gareth/ISL/) (Trevor Hastie)
- Package development:
    - [R packages](http://r-pkgs.had.co.nz/)

Learn: [R Course Finder](http://r-exercises.com/r-courses/)


## References and further reading

- [7 Tips For Getting Your Colleagues Hooked on R](http://scl.io/QZxZZl6u#gs.zMhz76Q)
- [Video: What is R?](https://www.youtube.com/embed/TR2bHSJ_eck)
- [How Companies Use R to Compete in a Data-Driven World](http://data-informed.com/companies-use-r-compete-data-driven-world/)
- [How the growth of R helps data-driven organizations succeed](http://www.slideshare.net/RevolutionAnalytics/how-the-growth-of-r-helps-datadriven-organizations-succeed)
- [A segmented model of CRAN package growth](https://www.r-bloggers.com/a-segmented-model-of-cran-package-growth/)
- [Coke vs Soda vs Pop : Linguistic trends analyzed with Twitter and R](https://www.r-bloggers.com/coke-vs-soda-vs-pop-linguistic-trends-analyzed-with-twitter-and-r/)